Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
edits intro

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
not large enough to prove termination for all terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

@@

\keywords
Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating Point Ranking Functions.

@@

When surveying the area of program termination chronologically, we observe an initial domination of  monolithic approaches based on a single measure shown to decrease
over all program paths %(syntactic characterisation of loops) 
\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by 
more recent techniques moving towards termination arguments based on Ramsey's theorem \cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}.
This proof style proves that a transition relation is disjunctively well founded by finding a set of simple
functions, each of which proves well foundedness for part of the transition relation.
The main justification for this paradigm shift lies in the simplicity of the local termination measures when compared to the global ones, e.g.
there are cases in which proofs based on local measures involve linear functions while corresponding global
measures require non-linear functions or lexicographic orders. 
%This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

One difficulty with a Ramsey-based approach is that validity of the
termination argument can only be checked by considering the \emph{transitive
closure} of the program, rather than a single step.  As such, there is
experimental evidence that most of the time is spent in reachability
analysis \cite{DBLP:conf/pldi/CookPR06}, requiring the support of powerful
safety checkers: there is a tradeoff between the complexity of the
termination arguments and that of checking their validity.

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments.  This
means that our ranking functions may involve non-linearity and lexicographic
orders (we do not commit to any particular syntactic form, and do not use
templates).

While the summary  in Figure~\ref{fig:handletable} supports the observation
that the majority of existing termination analyses are designed for linear
programs and linear ranking functions, it also identifies another common
assumption made by state of the art termination provers: that bit-vector
semantics and integer semantics give rise to the same termination behaviour. 
Thus, most of the techniques treat bit-vectors and floats as mathematical
integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

@@

%% %\noindent {\bf Floats vs. reals.} 
%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about floating point inherently difficult.
%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
%% While the former does not terminate for reals, but does for floats, the latter 
%% terminates for reals, but does not for floats.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
lexicographic, non-linear 
ranking functions supported by non-linear inductive invariants for loops with 
arbitrary guards and transitions over bit-vectors and floats.
It also computes non-linear recurrence sets, which comprise non-termination proofs.
Our encoding uniformly multiple loops and nested loops, without having to enumerate
lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

@@

\item We designed a bit-level accurate technique for computing ranking
functions that correctly accounts for the wrap-around behavior caused by
under- and overflows in bit-vector and floating point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

@@

Floating point arithmetic:

>>>>

\begin{abstract}
%
Proving program termination is typically done by finding a well-founded
\emph{ranking function} for the program states.  Existing termination
provers typically find ranking functions using either linear algebra or
templates.  As such they are often restricted to finding linear ranking
functions over mathematical integers.  This class of ranking functions is
insufficient for prove termination for many terminating programs, and
furthermore a termination argument for a program operating on mathematical
integers does not always lead to a termination argument for the same program
operating on fixed-width machine integers.

@@

\keywords
Termination, Program Synthesis, Lexicographic Ranking Functions, Bitvector Ranking Functions,
Floating-Point Ranking Functions.

@@

When surveying the area of program termination chronologically, we observe
an initial focus on monolithic approaches based on a single measure shown to
decrease over all program
paths~\cite{DBLP:conf/vmcai/P04,DBLP:conf/cav/BradleyMS05}, followed by more
recent techniques that use termination arguments based on Ramsey's
theorem~\cite{DBLP:conf/lpe/CodishG03,DBLP:conf/lics/PodelskiR04,DBLP:conf/pldi/CookPR06}. 
Briefly, the transition relation is shown to be disjunctively well founded
by finding a set of simple functions, each of which proves well foundedness
for part of the transition relation.  The main benefit of this approach is
the simplicity of local termination measures in contrast to global ones. 
For instance, there are cases in which linear arithmetic suffices when using
local measures, while corresponding global measures require non-linear
functions or lexicographic orders.

%This explains why approaches based on Ramsey's theorem are restricted to searching for linear termination arguments.

One drawback of the Ramsey-based approach is that the validity of the
termination argument relies on checking the \emph{transitive closure} of the
program, rather than a single step.  As such, there is experimental evidence
that most of the effort is spent in reachability
analysis~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/cav/KroeningSTW10},
requiring the support of powerful safety checkers: there is a tradeoff
between the complexity of the termination arguments and that of checking
their validity.

As Ramsey-based approaches are limited by the state of the art in safety
checking, recent research shifts back to more complex termination arguments
that are easier to
check~\cite{DBLP:conf/cav/KroeningSTW10,DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/HeizmannHP14}. 
Following the same trend, we investigate its extreme: \emph{unrestricted}
termination arguments.  This means that our ranking functions may involve
non-linearity and lexicographic orders: we do not commit to any particular
syntactic form, and do not use templates.

While the summary  in Figure~\ref{fig:handletable} supports the observation
that the majority of existing termination analyses are designed for linear
programs and linear ranking functions, it also identifies another common
assumption made by state-of-the-art termination provers: that bit-vector
semantics and integer semantics give rise to the same termination behaviour. 
Thus, most of the techniques treat bit-vectors and floats as mathematical
integers and reals,
respectively~\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

@@

%% %\noindent {\bf Floats vs. reals.} 
%% A scenario similar to the one for bit-vectors happens if floats were to be abstracted to unbounded reals by termination provers \cite{}. 
%% Consequently, these provers ignore potential rounding errors, under- and over-flows, which are precisely what makes  reasoning about
%% floating-point inherently difficult.
%% This approximation may lead to erroneous diagnosis of a program's terminating behaviour as illustrated by the loops in Figure~\ref{} and Figure~\ref{}.
%% While the former does not terminate for reals, but does for floats, the latter 
%% terminates for reals, but does not for floats.

@@

We propose a general framework that uniformly computes lexicographic,
non-linear ranking functions supported by non-linear inductive invariants
for loops with arbitrary guards and transitions over bit-vectors and floats. 
It also computes non-linear recurrence sets, which comprise non-termination
proofs.  Our encoding uniformly multiple loops and nested loops, without
having to enumerate lassos.
%
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

@@

\item We designed a bit-level accurate technique for computing ranking
functions that correctly accounts for the wrap-around behavior caused by
under- and overflows in bit-vector and floating-point arithmetic.  Our
technique is not restricted to finding linear ranking functions, but can
also compute lexicographic non-linear ones.

@@

Floating-point arithmetic:
